Single-seismometer Focal Mechanism and Uncertainty Estimation from Body Waves

### 2024-10-29 paper/main.tex
Up to algorithms

% import packages
\usepackage{gensymb, amssymb, nicefrac, bm, tikz}

@@

    [Suzan] Insert words here...\\
    Talk about inverse method for a highly nonlinear problem\\
    Writing tip: maximize information content of words

@@

    \subsection{Derivation}
    When a quake event happens, a single seismometer reads $P$, $SV$ and $SH$ body waves through
    which the respective amplitudes $A^P$, $A^{SV}$ and $A^{SH}$ along with their noise levels
    $\sigma_P$, $\sigma_{SV}$ and $\sigma_{SH}$ (representing standard deviation of an assumed normal
    distribution centered at the amplitude value) can be extracted.
    Given a ray path's azimuth $\phi$ and take-off angles $i$, $j$ of $P$ and $S$ waves respectively,
    we model the amplitudes as functions of the source mechanism's fault strike $\psi$, fault dip
    $\delta$ and slip rake $\lambda$, further accounting for the $P$- and $S$-velocities $\alpha_h$,
    $\beta_h$ at source depth $h$ \citep{AkiRichards1980}. The expressions are:

    \begin{align} \label{eq:1}
      \nonumber &A^P \sim \nicefrac{\left( s_R(3\cos^2(i) - 1) - q_R\sin(2i) - p_R\sin^2(i)\right)}{\alpha_h^3}\\
      &A^{SV} \sim \nicefrac{\left( 1.5 s_R\sin(2j) + q_R\cos(2j) + 0.5 p_R\sin(2j) \right)}{\beta_h^3}\\
      \nonumber &A^{SH} \sim \nicefrac{\left( q_L\cos(j) + p_L\sin(j) \right)}{\beta_h^3}
    \end{align}

@@

    \begin{align} \label{eq:2}
      \nonumber s_R &= 0.5 \sin(\lambda) \sin(2\delta)\\
      \nonumber q_R &= \sin(\lambda) \cos(2\delta) \sin(\psi_r) + \cos(\lambda) \cos(\delta) \cos(\psi_r)\\
      p_R &= \cos(\lambda) \sin(\delta) \sin(2\psi_r) - 0.5\sin(\lambda) \sin(2\delta) \cos(2\psi_r)\\
      \nonumber p_L &= 0.5\sin(\lambda) \sin(2\delta) \sin(2\psi_r) + \cos(\lambda) \sin(\delta) \cos(2\psi_r)\\
      \nonumber q_L &= -\cos(\lambda) \cos(\delta) \sin(\psi_r) + \sin(\lambda) \cos(2\delta) \cos(\psi_r)
    \end{align}

@@

    We thus need a metric against which to evaluate what the best and acceptable estimates are. The
    misfit function used by \citet{sita_potential_2022} is defined as the angle between the vector of
    observed amplitudes and that produced from a forward simulation of synthetic parameters.
    In this paper, we shall use a misfit $\mu$ of cosine similarity between the two vectors which has a
    1-to-1 correspondence with angle:

    \begin{align} \label{eq:3}
      \mu(\bm{p}) = \frac{\bm{A}_o^\top\bm{A}_s}{\|\bm{A}_o\|\|\bm{A}_s\|}
    \end{align}

@@

    \begin{align} \label{eq:4}
      \mu(\bm{p}) = \sqrt{\left( \Delta \frac{A^{P}}{A^{SV}}\right)^2 + \left( \Delta \frac{A^{SV}}{A^{SH}}\right)^2}
    \end{align}

    \begin{align} \label{eq:5}
      \mu(\bm{p}) = \sqrt{\log^2 \left( \Delta \frac{A^P}{A^{SV}} \right) + \log^2 \left( \Delta \frac{A^{SV}}{A^{SH}} \right)}
    \end{align}

@@

    \begin{align} \label{eq:6}
      &\mu(\bm{p}) \sim |\Sigma|^{-1/2} \exp \left( - \frac{1}{2} (\bm{A} - \bm{A}_o)^\top \bm{\Sigma}^{-1} (\bm{A} - \bm{A}_o) \right)
    \end{align}

    \subsection{Tolerance}
    Given vectors $\bm{A}_s$ and $\bm{A}_o$, we would like to know the largest cosine
    similarity for which $\bm{A}_s$ is considered an acceptable fit with respect to the recorded amplitudes
    while accounting for the asymmetry of error levels for different amplitudes. Visualized in 3D
    space, we have a confidence ellipsoid centered at $\bm{A}_o$ (equation \ref{eq:7}) such that
    acceptable fits are vectors which fall inside the elliptical cone tangent to it with vertex at the
    origin.
    \begin{align} \label{eq:7}
      E(\bm{p}) = (\bm{A} - \bm{A}_o)^\top \hat{\bm{\Sigma}}^{-1} (\bm{A} - \bm{A}_o)  = 1
    \end{align}

@@

    The vector $\bm{A} = (A^P, A^{SV}, A^{SH})^\top$ can be treated as a continuously differentiable nonlinear function of
    parameters $\bm{p} = (\psi, \delta, \lambda)^\top$, defined by equation \ref{eq:1}. We can thus apply
    directed search methods to minimize the following cost function:
    \begin{align} \label{eq:17}
      \mu(\bm{p}) = \hat{\bm{A}}_o^\top \hat{\bm{A}}
    \end{align}

    \textcolor{red}{Continue from here...}

@@

        \node (M) at (7,0) {$\mu$};

@@

        \draw (H1) -- (A3);
        \draw (H1) -- (A1);
        \draw (H1) -- (A2);
        \draw (H2) -- (A1);
        \draw (H2) -- (A2);
        \draw (H3) -- (A1);
        \draw (H3) -- (A2);
        \draw (H4) -- (A3);
        \draw (H5) -- (A3);
      \end{tikzpicture}
      \caption{Computational graph of forwayd model.}
      \label{fig:1}
    \end{figure}

@@

    Insert words here...

\section{Example: from Paula} \label{sec:example}

    Insert words here...

@@

    Insert words here...

@@

\section*{Data and code availability}
Zenodo, figshare, and Dryad to archive data and code. Citations for datasets and codes should be
included in the references, including citations for any seismic networks from which data was used.
Github is not considered a persistent repository, and we encourage authors to archive a snapshot of
any github-hosted code on zenodo.

>>>>

% import packages
\usepackage{amssymb, nicefrac, bm, tikz}

@@

    [Suzan] Insert words here...\\
    Writing tip: maximize information content of words

@@

  \subsection{Background}
    When a quake event happens, a single seismometer reads $P$, $SV$ and $SH$ body waves through
    which the respective amplitudes $A^P$, $A^{SV}$ and $A^{SH}$ along with their noise levels
    $\sigma_P$, $\sigma_{SV}$ and $\sigma_{SH}$ (representing standard deviation of an assumed normal
    distribution centered at the amplitude value) can be extracted.
    Given a ray path's azimuth $\phi$ and take-off angles $i$, $j$ of $P$ and $S$ waves respectively,
    we model the amplitudes as functions of the source mechanism's fault strike $\psi$, fault dip
    $\delta$ and slip rake $\lambda$, further accounting for the $P$- and $S$-velocities $\alpha_h$,
    $\beta_h$ at source depth $h$ \citep{AkiRichards1980}. The expressions are:

    \begin{align} \label{eq:1}
      \nonumber &A^P \sim \left( s_R(3\cos^2(i) - 1) - q_R\sin(2i) - p_R\sin^2(i)\right)/\alpha_h^3\\
      \nonumber &A^{SV} \sim \left( 1.5 s_R\sin(2j) + q_R\cos(2j) + 0.5 p_R\sin(2j) \right)/\beta_h^3\\
      &A^{SH} \sim \left( q_L\cos(j) + p_L\sin(j) \right)/\beta_h^3
    \end{align}

@@

    \begin{align} \label{eq:2}
      \nonumber s_R &= 0.5 \sin(\lambda) \sin(2\delta)\\
      \nonumber q_R &= \sin(\lambda) \cos(2\delta) \sin(\psi_r) + \cos(\lambda) \cos(\delta) \cos(\psi_r)\\
      \nonumber p_R &= \cos(\lambda) \sin(\delta) \sin(2\psi_r) - 0.5\sin(\lambda) \sin(2\delta) \cos(2\psi_r)\\
      \nonumber p_L &= 0.5\sin(\lambda) \sin(2\delta) \sin(2\psi_r) + \cos(\lambda) \sin(\delta) \cos(2\psi_r)\\
      q_L &= -\cos(\lambda) \cos(\delta) \sin(\psi_r) + \sin(\lambda) \cos(2\delta) \cos(\psi_r)
    \end{align}

@@

    We thus need a metric against which to evaluate what the best and acceptable estimates are. The
    misfit function used by \citet{sita_potential_2022} is defined as the angle between the vector of
    observed amplitudes and that produced from a forward simulation of synthetic parameters.
    In this paper, we shall use a misfit $\Phi$ of negative cosine similarity between the two vectors
    which has a 1-to-1 correspondence with angle:

    \begin{align} \label{eq:3}
      \Phi(\bm{p}) = \frac{-\bm{A}_o^\top\bm{A}_s}{\|\bm{A}_o\|\|\bm{A}_s\|}
    \end{align}

@@

    \begin{align} \label{eq:4}
      \Phi(\bm{p}) = \sqrt{\left( \Delta \frac{A^{P}}{A^{SV}}\right)^2 + \left( \Delta \frac{A^{SV}}{A^{SH}}\right)^2}
    \end{align}

    \begin{align} \label{eq:5}
      \Phi(\bm{p}) = \sqrt{\log^2 \left( \Delta \frac{A^P}{A^{SV}} \right) + \log^2 \left( \Delta \frac{A^{SV}}{A^{SH}} \right)}
    \end{align}

@@

    \begin{align} \label{eq:6}
      &\Phi(\bm{p}) \sim |\Sigma|^{-1/2} \exp \left( - \frac{1}{2} (\bm{A} - \bm{A}_o)^\top \bm{\Sigma}^{-1} (\bm{A} - \bm{A}_o) \right)
    \end{align}

  \subsection{Tolerance derivation}
    Given vectors $\bm{A}_s$ and $\bm{A}_o$, we would like to know the largest cosine
    similarity for which $\bm{A}_s$ is considered an acceptable fit with respect to the recorded amplitudes
    while accounting for the asymmetry of error levels for different amplitudes. Visualized in 3D
    space, we have a confidence ellipsoid centered at $\bm{A}_o$ (equation \ref{eq:7}) such that
    acceptable fits are vectors which fall inside the elliptical cone tangent to it with vertex at the
    origin.
    \begin{align} \label{eq:7}
      E(\bm{p}) = (\bm{A} - \bm{A}_o)^\top \hat{\bm{\Sigma}}^{-1} (\bm{A} - \bm{A}_o)  = 1
    \end{align}

@@

  \subsection{Misfit gradient}
    The vector $\bm{A} = (A^P, A^{SV}, A^{SH})^\top$ can be treated as a continuously differentiable 
    nonlinear function of parameters $\bm{p} = (\psi, \delta, \lambda)^\top$, defined by equation 
    \ref{eq:1}. We can thus apply directed search methods to minimize the following cost function:
    \begin{align} \label{eq:17}
      \Phi(\bm{p}) = -\hat{\bm{A}}_o^\top \hat{\bm{A}}
    \end{align}

    To calculate the gradient, first note that $\bm{A}$ is a linear combination of nonlinear functions
    of $\bm{p}$ (equation \ref{eq:2}).

@@

        \node (M) at (7,0) {$\Phi$};

@@

        \draw (H1) -- (A3);
        \draw (H1) -- (A1);
        \draw (H1) -- (A2);
        \draw (H2) -- (A1);
        \draw (H2) -- (A2);
        \draw (H3) -- (A1);
        \draw (H3) -- (A2);
        \draw (H4) -- (A3);
        \draw (H5) -- (A3);
      \end{tikzpicture}
      \caption{Computational graph of forward model (edges are non-zero partial derivatives).}
      \label{fig:1}
    \end{figure}

    Define a vector $\bm{\eta} = (s_R, q_R, p_R, q_L, p_L)^\top$. Then we can encode
    the misfit as
    \begin{align} \label{eq:18}
      \Phi(\bm{p}) = \Phi(\bm{A}(\bm{\eta}(\bm{p})))
    \end{align}

    \noindent This composition is visualized as a computational graph in figure \ref{fig:1}.
    Each stage of the composition has an associated Jacobian, which we can demonstrate layer by layer:

    i) From $\bm{p}$ to $\bm{\eta}$, we can use equation \ref{eq:2} to define the Jacobian
    as follows:
    \begin{align*}
      &a = \sin(\lambda)\;\;\;\;\;\;\;\;\;\;\;\;\; b = \cos(2\delta)\;\;\;\;\;\;\;\;\;\;\; c = \cos(\lambda)\\
      &d = \sin(2\delta)\;\;\;\;\;\;\;\;\;\;\;\; e = \sin(\delta)\;\;\;\;\;\;\;\;\;\;\;\; f = \cos(\delta)\\
      &g = \sin(\psi - \phi)\;\;\;\;\;\; h = \cos(\psi - \phi)\;\;\;\;\; p = cos(2(\psi - \phi))\\
      &q = \sin(2(\psi - \phi))
    \end{align*}
    \begin{align} \label{eq:19}
      \nonumber&\bm{J_p}(\bm{\eta}) = \\
      &\begin{bmatrix}
        0         &ab         &0.5cd\\
        abh-cfg   &-2adg-ceh  &cbg-afh\\
        2cep+adq  &cfq-abp    &-aeq-0.5cdp\\
        -cfh-abg  &ceg-2adh   &afg+cbh\\
        adp-2ceq  &abq+cfp    &0.5cdq-aep
      \end{bmatrix}
    \end{align}

    ii) From $\bm{\eta}$ to $A$, we have a linear transformation whose Jacobian is the matrix
    of coefficients from equation \ref{eq:1}:
    \begin{align} \label{eq:20}
      \nonumber&\bm{J_\eta}(\bm{A}) = \\
      &\begin{bmatrix}
        \frac{3\cos^2(i)-1}{\alpha^3}    &\frac{-\sin(2i)}{\alpha^3}    &\frac{-sin^2(i)}{\alpha^3}   &0    &0\\
        \frac{1.5\sin(2j)}{\beta^3}      &\frac{\cos(2j)}{\beta^3}      &\frac{0.5\sin(2j)}{\beta^3}  &0    &0\\
        0         &0           &0        &\frac{\cos(j)}{\beta^3}       &\frac{\sin(j)}{\beta^3}
      \end{bmatrix}
    \end{align}

    iii) From $\bm{A}$ to $\Phi$, we have a scalar function whose Jacobian is the gradient
    transposed.
    \begin{align} \label{eq:21}
      \nonumber\nabla_{\bm{A}}\Phi &= \frac{\|\bm{A}\|\nabla_{\bm{A}}(-\hat{\bm{A}}_o^\top\bm{A}) + 
      \hat{\bm{A}}_o^\top \bm{A} \nabla_{\bm{A}}(\|\bm{A}\|)}{\|\bm{A}\|^2}\\
            &= \frac{1}{\|\bm{A}\|}(\Phi\hat{\bm{A}} - \hat{\bm{A}}_o)
    \end{align}

    Combining \ref{eq:19}, \ref{eq:20} and \ref{eq:21} gives the overall gradient:
    \begin{align} \label{eq:22}
      \nabla_{\bm{p}}\Phi = \bm{J_p}(\bm{\eta})^\top \bm{J_\eta}(\bm{A})^\top \nabla_{\bm{A}}\Phi
    \end{align}

  \subsection{Optimization algorithms}
    Something about gradient-based methods. Try different algorithms, they may produce
    interesting results >> IEMS 450-2.

    \noindent \textbf{TODO:} Create one 3D image with the results of gradient descent from random points in space.

@@

    \subsection{Error propagation}
      Say something here, composed Jacobian is important >> Earth 353, IEMS 401.

      \noindent Emphasize the interest in modality.

    \subsection{Monte Carlo methods}
      Place to talk about different seeding stategies and their effect on uncertainty.
      Why hybrid search was used, why systematic over random.

    \subsection{Kagan angle}
      An alternative way of quantifying model deviations.

\section{Testing} \label{sec:example}

    \subsection{Synthetic data}
      Say something here about the Monte Carlo

    \subsection{Earth data}
      From Uganda (Paula)

    \subsection{Mars data}
      From Insight (Suzan/Maddy)

@@

    \subsection{Performance overview}
      Put some words here...

    \subsection{Growth avenues}
      Put some more words here...

@@

\section*{Data and code availability}
Create a new repo with beautiful code, READMEs and everything. One function/algorithm at a time.
